---
title: "🎨 aRtpalette Package"
format: html
author: "Khoi Nguyen"
execute:
  message: false
  warning: false
---

## Overview

`aRtpalette` is an R package that brings art-inspired color palettes into
data visualizations. The package lets you store palettes as structured objects,
preview them as color swatches, and expand them when a chart needs more colors
than the source artwork provides.

The inspiration came from the frustration of manually copying hex codes from
paintings into ggplot2 charts with no clean workflow. I wanted a tool that
treated color as data — something you could store, validate, visualize, and
manipulate programmatically.

---

## Installation

```r
# Option 1: using devtools
devtools::install_github("ADC-405-S26/aRtpalette")

# Option 2: using pak
pak::pak("ADC-405-S26/aRtpalette")
```

---

## Built-in Dataset

The package includes `artwork_palettes` — a table of 20 color palettes
extracted from iconic artworks across art history. Each row is one painting,
with five hex color codes and metadata including artist, title, and art movement.

```{r}
library(aRtpalette)
library(ggplot2)
data(artwork_palettes)
head(artwork_palettes[, c("palette_name", "artist", "style", "hex1", "hex2", "hex3")])
```

---

## Core Functions

### `palette_from_hex()` — store colors as a validated object

Validates every hex code on input and returns a clear error message if any
is malformed — so problems surface immediately rather than silently breaking
a chart later.

```{r}
row <- artwork_palettes[artwork_palettes$palette_name == "starry_night", ]

starry_night <- palette_from_hex(
  hex_codes = c(row$hex1, row$hex2, row$hex3, row$hex4, row$hex5),
  name      = "starry_night"
)

starry_night
```

---

### `plot_palette()` — preview before you apply

Seeing the palette as swatches before applying it to real data saves the
back-and-forth of guessing whether colors will work together.

```{r fig.height=2}
plot_palette(starry_night)
```

---

### `apply_palette()` — expand for any chart size

If you ask for more colors than the palette contains, the function interpolates
between existing colors — letting you expand any palette creatively for any
chart size.

```{r}
# Exactly 3 colors for a 3-group chart
apply_palette(starry_night, n = 3)
```

```{r}
# Expand to 8 colors via interpolation
apply_palette(starry_night, n = 8)
```

---

## Full Workflow Example

All three functions working together in one clean pipeline — from dataset
to validated palette to preview to chart, with no hex codes typed by hand.

```{r fig.height=2}
# Step 1 — pull the Starry Night row and turn it into a validated palette object
row <- artwork_palettes[artwork_palettes$palette_name == "starry_night", ]

pal <- palette_from_hex(
  c(row$hex1, row$hex2, row$hex3, row$hex4, row$hex5),
  name = "starry_night"
)

# Step 2 — preview the palette before using it
plot_palette(pal)
```

```{r fig.height=4}
# Step 3 — count artworks by art style from the built-in dataset
style_counts <- as.data.frame(table(artwork_palettes$style))
colnames(style_counts) <- c("style", "count")

# Step 4 — expand palette to exactly the number of style groups needed
chart_colors <- apply_palette(pal, n = nrow(style_counts))

# Step 5 — apply directly to ggplot2
ggplot(style_counts, aes(x = reorder(style, count),
                          y = count,
                          fill = style)) +
  geom_bar(stat = "identity", width = 0.7) +
  scale_fill_manual(values = chart_colors) +
  coord_flip() +
  labs(
    title    = "Number of Artworks by Art Style",
    subtitle = "Colors from Van Gogh's The Starry Night",
    x        = "Art Style",
    y        = "Count"
  ) +
  theme_classic() +
  theme(legend.position = "none")
```

---

## Summary

`aRtpalette` provides three functions that work together as a pipeline:

- **`palette_from_hex()`** — store hex codes as a validated, named palette object
- **`plot_palette()`** — preview the palette as color swatches before applying
- **`apply_palette()`** — extract or expand the palette to exactly the colors your chart needs

Combined with the built-in `artwork_palettes` dataset of 20 palettes from
iconic artworks, the package makes it easy to bring art history into R without
any manual hex code lookup.

---

## Learn More

See the full vignette for a detailed walkthrough:

```r
vignette("getting-started-with-aRtpalette")
```

## Source

- GitHub: [ADC-405-S26/aRtpalette](https://github.com/ADC-405-S26/aRtpalette)