---
title: "🏎️ F1 SQL Analysis"
author: "Khoi Nguyen & Tiana Ho"
format: html
---

## Overview

We used the Formula 1 dataset from Kaggle to explore how performance patterns 
have changed over time — looking at driver success, team dominance, grid 
advantage, international diversity, and scoring trends.

---

## Setup & Connection
```{r setup}
#| message: false
#| warning: false
#| include: false
library(tidyverse)
library(DBI)
library(duckdb)
library(ggrepel)
```

```{r}
con_f1 <- dbConnect(duckdb::duckdb(), dbdir = "f1_db")
```

---

## Loading the Data

We read in three CSV files — races, drivers, and results — which form the 
core of the relational database.

```{r}
races   <- readr::read_csv("races.csv")
drivers <- readr::read_csv("drivers.csv")
results <- readr::read_csv("results.csv")
```

---

## Building the Database

We created three relational tables in DuckDB and loaded each CSV into them.

```{sql connection=con_f1}
DROP TABLE IF EXISTS races;
```
```{sql connection=con_f1}
CREATE TABLE races (
  raceId INTEGER, year INTEGER, round INTEGER, circuitId INTEGER,
  name VARCHAR, date VARCHAR, time VARCHAR, url VARCHAR,
  fp1_date VARCHAR, fp1_time VARCHAR, fp2_date VARCHAR, fp2_time VARCHAR,
  fp3_date VARCHAR, fp3_time VARCHAR, quali_date VARCHAR, quali_time VARCHAR,
  sprint_date VARCHAR, sprint_time VARCHAR);
COPY races FROM 'races.csv' (FORMAT CSV, HEADER, NULL '\N');
```

```{sql connection=con_f1}
DROP TABLE IF EXISTS drivers;
```
```{sql connection=con_f1}
CREATE TABLE drivers (
  driverId INTEGER, driverRef VARCHAR, number VARCHAR, code VARCHAR,
  forename VARCHAR, surname VARCHAR, dob VARCHAR, nationality VARCHAR, url VARCHAR);
COPY drivers FROM 'drivers.csv' (FORMAT CSV, HEADER, NULL '\N');
```

```{sql connection=con_f1}
DROP TABLE IF EXISTS results;
```
```{sql connection=con_f1}
CREATE TABLE results (
  resultId INTEGER, raceId INTEGER, driverId INTEGER, constructorId INTEGER,
  number INTEGER, grid INTEGER, position VARCHAR, positionText VARCHAR,
  positionOrder INTEGER, points DOUBLE, laps INTEGER, time VARCHAR,
  milliseconds VARCHAR, fastestLap VARCHAR, rank VARCHAR,
  fastestLapTime VARCHAR, fastestLapSpeed VARCHAR, statusId INTEGER);
COPY results FROM 'results.csv' (FORMAT CSV, HEADER, NULL '\N');
```

---

## Queries

### Q1 — All-Time Driver Rankings
Who scored the most career points? We joined results with drivers and 
summed points across all races.

```{sql connection=con_f1}
SELECT
  d.forename || ' ' || d.surname AS driver_name,
  d.nationality,
  COUNT(r.raceId) AS total_races,
  SUM(r.points) AS total_points,
  MAX(r.points) AS best_race_points
FROM results AS r
JOIN drivers AS d ON r.driverId = d.driverId
WHERE r.points IS NOT NULL
GROUP BY d.driverRef, d.forename, d.surname, d.nationality
ORDER BY total_points DESC
LIMIT 10;
```

### Q2 — Constructor Dominance by Decade
How did team success shift across eras? We bucketed results by decade 
and summed points per constructor.

```{sql connection=con_f1}
SELECT
  res.constructorId,
  CASE
    WHEN ra.year BETWEEN 1950 AND 1959 THEN '1950s'
    WHEN ra.year BETWEEN 1960 AND 1969 THEN '1960s'
    WHEN ra.year BETWEEN 1970 AND 1979 THEN '1970s'
    WHEN ra.year BETWEEN 1980 AND 1989 THEN '1980s'
    WHEN ra.year BETWEEN 1990 AND 1999 THEN '1990s'
    WHEN ra.year BETWEEN 2000 AND 2009 THEN '2000s'
    WHEN ra.year BETWEEN 2010 AND 2019 THEN '2010s'
    ELSE '2020s'
  END AS decade,
  SUM(res.points) AS total_points,
  COUNT(DISTINCT res.driverId) AS drivers_used
FROM results AS res
JOIN races AS ra ON res.raceId = ra.raceId
WHERE res.points IS NOT NULL
GROUP BY res.constructorId, decade
ORDER BY decade, total_points DESC;
```

### Q3 — Win Rate by Driver
Which drivers converted starts into wins most efficiently? 
Minimum 50 starts required.

```{sql connection=con_f1}
SELECT
  d.forename || ' ' || d.surname AS driver_name,
  d.nationality,
  COUNT(*) AS total_starts,
  SUM(CASE WHEN res.positionOrder = 1 THEN 1 ELSE 0 END) AS wins,
  ROUND(100.0 * SUM(CASE WHEN res.positionOrder = 1 THEN 1 ELSE 0 END) / COUNT(*), 2) AS win_rate_pct
FROM results AS res
JOIN drivers AS d ON res.driverId = d.driverId
GROUP BY d.driverRef, d.forename, d.surname, d.nationality
HAVING COUNT(*) >= 50
ORDER BY win_rate_pct DESC
LIMIT 15;
```

### Q4 — Grid Position vs. Finish
Does starting at the front guarantee a strong result?

```{sql connection=con_f1}
SELECT
  res.grid AS grid_position,
  COUNT(*) AS total_occurrences,
  ROUND(AVG(res.positionOrder), 2) AS avg_finish_position,
  ROUND(AVG(res.points), 2) AS avg_points_earned
FROM results AS res
WHERE res.grid BETWEEN 1 AND 20
  AND res.positionOrder BETWEEN 1 AND 20
GROUP BY res.grid
ORDER BY grid_position;
```

### Q5 — International Diversity Over Time
How global did F1 become across different eras?

```{sql connection=con_f1}
SELECT
  CASE
    WHEN ra.year BETWEEN 1950 AND 1969 THEN 'Early Era (1950-1969)'
    WHEN ra.year BETWEEN 1970 AND 1989 THEN 'Mid Era (1970-1989)'
    WHEN ra.year BETWEEN 1990 AND 2009 THEN 'Modern Era (1990-2009)'
    ELSE 'Hybrid Era (2010+)'
  END AS era,
  COUNT(DISTINCT d.nationality) AS unique_nationalities,
  COUNT(DISTINCT d.driverId) AS unique_drivers,
  COUNT(DISTINCT ra.raceId) AS total_races
FROM results AS res
JOIN drivers AS d ON res.driverId = d.driverId
JOIN races AS ra ON res.raceId = ra.raceId
GROUP BY era
ORDER BY MIN(ra.year);
```

### Q6 — Wins by Nationality
Which countries produced the most race winners?

```{sql connection=con_f1}
SELECT
  d.nationality,
  COUNT(DISTINCT d.driverId) AS total_drivers,
  SUM(CASE WHEN res.positionOrder = 1 THEN 1 ELSE 0 END) AS total_wins,
  SUM(res.points) AS total_points
FROM results AS res
JOIN drivers AS d ON res.driverId = d.driverId
WHERE res.points IS NOT NULL
GROUP BY d.nationality
HAVING SUM(CASE WHEN res.positionOrder = 1 THEN 1 ELSE 0 END) > 0
ORDER BY total_wins DESC
LIMIT 12;
```

### Q7 — Podium Rate by Driver
Beyond wins, who stayed on the podium most consistently? 
Minimum 30 starts required.

```{sql connection=con_f1}
SELECT
  d.forename || ' ' || d.surname AS driver_name,
  d.nationality,
  COUNT(*) AS total_starts,
  SUM(CASE WHEN res.positionOrder <= 3 THEN 1 ELSE 0 END) AS podiums,
  ROUND(100.0 * SUM(CASE WHEN res.positionOrder <= 3 THEN 1 ELSE 0 END) / COUNT(*), 1) AS podium_rate_pct
FROM results AS res
JOIN drivers AS d ON res.driverId = d.driverId
GROUP BY d.driverRef, d.forename, d.surname, d.nationality
HAVING COUNT(*) >= 30
ORDER BY podium_rate_pct DESC
LIMIT 12;
```

### Q8 — Scoring Trends Over Time
How did rule changes affect points per season?

```{sql connection=con_f1}
SELECT
  ra.year AS season_year,
  COUNT(DISTINCT ra.raceId) AS races_in_season,
  ROUND(AVG(res.points), 3) AS avg_points_per_result,
  MAX(res.points) AS max_points_in_race
FROM results AS res
JOIN races AS ra ON res.raceId = ra.raceId
WHERE res.points IS NOT NULL AND res.points > 0
GROUP BY ra.year
ORDER BY season_year;
```

---

## Visualizations

### Career Points vs. Win Rate
Top drivers plotted by total career points and win percentage. 
Bubble size reflects number of starts.

```{r}
q1 <- dbGetQuery(con_f1, "
  SELECT d.forename || ' ' || d.surname AS driver_name, d.nationality,
    COUNT(r.raceId) AS total_races, SUM(r.points) AS total_points,
    MAX(r.points) AS best_race_points
  FROM results AS r JOIN drivers AS d ON r.driverId = d.driverId
  WHERE r.points IS NOT NULL
  GROUP BY d.driverRef, d.forename, d.surname, d.nationality
  ORDER BY total_points DESC LIMIT 10")

q3 <- dbGetQuery(con_f1, "
  SELECT d.forename || ' ' || d.surname AS driver_name, d.nationality,
    COUNT(*) AS total_starts,
    SUM(CASE WHEN CAST(res.positionOrder AS INT) = 1 THEN 1 ELSE 0 END) AS wins,
    ROUND(100.0 * SUM(CASE WHEN CAST(res.positionOrder AS INT) = 1 THEN 1 ELSE 0 END) / COUNT(*), 2) AS win_rate_pct
  FROM results AS res JOIN drivers AS d ON res.driverId = d.driverId
  GROUP BY d.driverRef, d.forename, d.surname, d.nationality
  HAVING COUNT(*) >= 50 ORDER BY win_rate_pct DESC LIMIT 15")

q4 <- dbGetQuery(con_f1, "
  SELECT res.grid AS grid_position, COUNT(*) AS total_occurrences,
    ROUND(AVG(res.positionOrder), 2) AS avg_finish_position,
    ROUND(AVG(res.points), 2) AS avg_points_earned
  FROM results AS res
  WHERE res.grid BETWEEN 1 AND 20 AND res.positionOrder BETWEEN 1 AND 20
  GROUP BY res.grid ORDER BY grid_position")

q6 <- dbGetQuery(con_f1, "
  SELECT d.nationality, COUNT(DISTINCT d.driverId) AS total_drivers,
    SUM(CASE WHEN CAST(res.positionOrder AS INT) = 1 THEN 1 ELSE 0 END) AS total_wins,
    SUM(res.points) AS total_points
  FROM results AS res JOIN drivers AS d ON res.driverId = d.driverId
  WHERE res.points IS NOT NULL GROUP BY d.nationality
  HAVING total_wins > 0 ORDER BY total_wins DESC LIMIT 12")

q8 <- dbGetQuery(con_f1, "
  SELECT CAST(ra.year AS INT) AS season_year,
    COUNT(DISTINCT ra.raceId) AS races_in_season,
    ROUND(AVG(res.points), 3) AS avg_points_per_result,
    MAX(res.points) AS max_points_in_race
  FROM results AS res JOIN races AS ra ON res.raceId = ra.raceId
  WHERE res.points IS NOT NULL AND res.points > 0
  GROUP BY CAST(ra.year AS INT) ORDER BY season_year")
```

```{r}
#| fig-width: 10
#| fig-height: 7
viz1_data <- q1 %>%
  left_join(q3 %>% select(driver_name, win_rate_pct, total_starts), by = "driver_name") %>%
  filter(!is.na(win_rate_pct))

ggplot(viz1_data, aes(x = total_points, y = win_rate_pct,
  size = total_races, color = nationality, label = driver_name)) +
  geom_point(alpha = 0.8) +
  geom_text_repel(size = 3.2, fontface = "bold", max.overlaps = 15) +
  scale_size_continuous(range = c(4, 14), name = "Total Races") +
  scale_color_brewer(palette = "Set1", name = "Nationality") +
  labs(title = "F1 All-Time Greats: Career Points vs. Win Rate",
    subtitle = "Bubble size = number of race starts | Only drivers with 50+ starts shown",
    x = "Total Career Points", y = "Win Rate (%)",
    caption = "Source: Kaggle Formula 1 World Championship Dataset") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(color = "grey40"),
    panel.grid.minor = element_blank())
```

### Grid Position vs. Finishing Position
Starting at the front strongly predicts finishing near the front.

```{r}
#| fig-width: 9
#| fig-height: 6
ggplot(q4, aes(x = grid_position, y = avg_finish_position)) +
  geom_line(color = "#E10600", linewidth = 1.2) +
  geom_point(aes(size = avg_points_earned), color = "#E10600", alpha = 0.85) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey60") +
  annotate("text", x = 16, y = 14, label = "Perfect prediction\n(grid = finish)",
    color = "grey50", size = 3.2, fontface = "italic") +
  scale_size_continuous(range = c(2, 9), name = "Avg Points Earned") +
  scale_x_continuous(breaks = 1:20) + scale_y_continuous(breaks = 1:20) +
  labs(title = "Does Starting Position Decide the Race?",
    subtitle = "Average finishing position by grid slot | Point size = average points scored",
    x = "Grid (Starting) Position", y = "Average Finishing Position",
    caption = "Source: Kaggle Formula 1 World Championship Dataset") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(color = "grey40"),
    panel.grid.minor = element_blank())
```

### Wins by Nationality
British and German drivers lead all-time race wins by country.

```{r}
#| fig-width: 9
#| fig-height: 6
q6 %>%
  mutate(nationality = fct_reorder(nationality, total_wins)) %>%
  ggplot(aes(x = nationality, y = total_wins)) +
  geom_col(width = 0.75, fill = "#E10600") +
  geom_text(aes(label = total_wins), hjust = -0.2, size = 3.8, fontface = "bold") +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(title = "Which Countries Produce the Most F1 Winners?",
    subtitle = "Total race wins by driver nationality — all time",
    x = NULL, y = "Total Race Wins",
    caption = "Source: Kaggle Formula 1 World Championship Dataset") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(color = "grey40"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank())
```

### F1 Scoring Trends Over Time
The 2010 points system change created a clear jump in average points per race.

```{r}
#| fig-width: 11
#| fig-height: 6
ggplot(q8, aes(x = season_year, y = avg_points_per_result)) +
  geom_line(color = "#E10600", linewidth = 1.1) +
  geom_point(color = "#E10600", size = 1.8) +
  geom_vline(xintercept = 2010, linetype = "dashed", color = "grey40") +
  annotate("text", x = 2011, y = max(q8$avg_points_per_result) * 0.9,
    label = "New points\nsystem (2010)",
    color = "grey30", size = 3.2, hjust = 0, fontface = "italic") +
  scale_x_continuous(breaks = seq(1950, 2020, 10)) +
  labs(title = "F1 Scoring Landscape Over Time",
    subtitle = "Average points per race entry per season (points-scoring finishes only)",
    x = "Season", y = "Average Points Per Race Result",
    caption = "Source: Kaggle Formula 1 World Championship Dataset") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(color = "grey40"),
    panel.grid.minor = element_blank())
```

---

## Sources
- Kaggle F1 Dataset: https://www.kaggle.com/datasets/rohanrao/formula-1-world-championship-1950-2020
- DuckDB R docs: https://duckdb.org/docs/api/r
- ggplot2: https://ggplot2.tidyverse.org

```{r}
DBI::dbDisconnect(con_f1, shutdown = TRUE)
```